How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-12-09 analysis/paper/paper.Rmd
added affliation and edit 7D to 9D

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene. This is an important period because new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments. Groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviors. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. Population increased before the appearance of stemmed points.

@@

Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)). Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009). The MAT of MIS 2, including LGM, was relatively stable. The temperature increased again from late MIS 2 towards MIS 1. Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer. The range of MAT at the sites over our study period is 2-10℃. Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃). The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka. The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka. Figure 7D shows a negative relationship between temperature and elevation. For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level. Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.

>>>>

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park (University of Washington)
  - Ben Marwick (University of Washington)
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene. This is an important period because new tools, such as stemmed points and blades, appeared in the archaeological record. We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials. We explore environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models. We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments. Groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviors. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. Population increased before the appearance of stemmed points.

@@

Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)). Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009). The MAT of MIS 2, including LGM, was relatively stable. The temperature increased again from late MIS 2 towards MIS 1. Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer. The range of MAT at the sites over our study period is 2-10℃. Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃). The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka. The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka. Figure 9D shows a negative relationship between temperature and elevation. For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level. Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.
